---
title: "Reef Check Invert Data"
author: "Hannah Hayes"
format: html
editor: visual
---

## The Objective

The objective of this quarto document is to use the ReefCheck data from Monterey to understand the density of urchins m^2^. So we can use this data to inform the density of urchins we should use in the actual experiment.

## Load Packages

```{r}
#any packages that haven't been installed install

library(tidyverse)
library(ggplot2)
library(dplyr)

```

## Read in Data

```{r}
rc <- read.csv("Invert_California_raw_2024_ReefCheck.csv") #read in data and store as rc
head(rc)
```

## Tidy Data 
We want to select for the information we are interested in. We want to narrow it down to only urchins, past Point Concepcion and before Mendocino, and data from 2024.

```{r}
rcurch <- rc %>% #storing this as a new data frame and opening up the pipeline
  mutate(Date = mdy(Date)) %>%  #changing the date format from MM/DD/YY
  filter(Year == 2024) %>%  #filtering columns for data from year 2024
  filter(Classcode %in% c("Purple Urchin", "Red Urchin")) %>%  #filter for only red and purple urchins
  filter(between(Latitude,34.45,36.6)) #filter for sites past point concepcion and ending in Monterey

head(rcurch) #check to see if the formatting is correct
```

## Visualize Red vs Purple Urchins

Awesome! Now we have a data set that has only purple and red urchins. Now let's visualize the data to help us understand how abundant red urchins are in this area.

### Scatterplot
```{r}
rvp <- rcurch %>% #opening data frame and storing plot as rvp 
  ggplot(aes(x = Latitude , y = Amount , color = Classcode)) + #defining our variables
  geom_point() #This adds the data into the plot 
rvp #printing the plot 
```

This plot looks okay, but let's try making this into a histogram and a bar plot 

### Barplot
```{r}
rvpbar <- rcurch %>% 
  ggplot(aes(x = Site, y = Amount, fill = Classcode)) +
           geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
rvpbar
```



